---
title: "Chapter 4 exercises"
date: "1/29/2017"
output: html_document
---

```{r, message = F, warning = F}
suppressWarnings(library(dplyr))
```


## Question 1

Model: $log(weight) = -3.5 + 2 * log(height) + error$

Weight in pounds. height in inches. Error sd = 0.25.

### a) Fill in the blanks (bolded): 

Approximately 65% of the persons will have weights within a factor of __`r round(exp(-0.25), 2)`__ and __`r round(exp(0.25), 2)`__ of their predicted values. 

Answered using the error sd: ($e^{-0.25}$, $e^{0.25}$)

### b) Plot a scatterplot and regression line for this model.

```{r}
set.seed(10)

# make up some some simple height input data
height = rnorm(1000, 68, 3)

log_weight = -3.5 + 2 * log(height) + rnorm(1000, 0, 0.25)

q1_model = lm(log_weight ~ log(height))

summary(q1_model)

plot(log(height), log_weight)
curve(cbind(1, x) %*% coef(q1_model), add = T, col = 3)

```

## Question 2

Use earnings data: earnings, sex, height, weight

### a) Load and QC the data

There is no weight variable, so I pulled in birth year.

*load:
```{r}
earnings_dat = dplyr::select(foreign::read.dta("ARM_Data/earnings/heights.dta"),
               earn, sex, height, yearbn)   

str(earnings_dat)
```

*check the variables

```{r}
par(mfrow = c(2,2))
hist(earnings_dat$earn, breaks = 100)
hist(earnings_dat$sex, breaks = 100)
hist(earnings_dat$height, breaks = 100)
hist(earnings_dat$yearbn, breaks = 100) #

par(mfrow = c(1,1))
ftable(cut(earnings_dat$earn, breaks = c(0, 1, 10, Inf), right = F))

```

Earnings has a bunch of 0s and NAs. 0s are odd so I will remove them, IRL this should be investigated further. There is a bump in yearbn at 99, probably a missing code so will remove those too.

*complete case for covenience, make final variables
```{r}
earnings_clean = subset(earnings_dat, !is.na(earn) & earn > 0 & yearbn < 99)

#stole these from other person posting answers
earnings_clean$earn = earnings_clean$earn/1000
earnings_clean$sex = factor(earnings_clean$sex, labels=c("male", "female"),
                            ordered = F)

#note that once you center birth year that the interpretation is the same as decreasing age, so flip the sign to represent getting older. See below.
earnings_clean$yearbn_ctr =  earnings_clean$yearbn - mean(earnings_clean$yearbn)
earnings_clean$age_yrs_ctr = mean(earnings_clean$yearbn) - earnings_clean$yearbn
mean(earnings_clean$yearbn)

```

Age$_i$ = current_year - $yearbn_i$

Age_center$_i$ = Age$_i$ - E[Age]    
Age_center$_i$ = current_year - $yearbn_i$ - E[current_year - $yearbn$]    
Age_center$_i$ = E[$yearbn$] - $yearbn_i$. 

The range of this variable is 72 years, which means it could range from like 16 to 89.

I chose to the yearbn centered variable (average year 1949) because I think the interpretation is actually clearer than ages without a reference year.

### b) Run model: earnings ~ height. Transform the data so that the intercept is the average earnings at the average height.

Center height at mean. **Intercept earnings move from -62K to 23K with transformation.** Note how the rest of the model is unaffected.

```{r}
earn_model2b = lm(earn ~ height, data = earnings_clean)
summary(earn_model2b)


earnings_clean$height_ctr = earnings_clean$height - mean(earnings_clean$height)
  
  
earn_model2b_2 = lm(earn ~ height_ctr, data = earnings_clean)
summary(earn_model2b_2)


```

### c) Fit some models

Will try transforming earnings because of the skew. I normalized height ignoring issues with sex. Year born was scaled in decade increments.

Trying the following models (set up *a prior*):
1. All three predictors (height, sex, yearbn): with log earn and without to test log transform diagnostics.    
2. Sex and yearbn interaction.    
3. Sex and height interaction: will include sex*yearbn depending on results.

Note that mean height is 67 inches and mean year born is 1949. Earnings is scaled by $1000.

Models 1:    
The diagnostics for this model are not great, there seems to be some clusters.
```{r}
# earnings_clean = earnings_clean %>% group_by(sex) %>%
#   mutate(mean_height = mean(height), sd_height = sd(height)) %>%
#   ungroup() %>%
#   rowwise() %>% mutate(height_norm_sex = (height - mean_height)/sd_height)

earnings_clean$log_earn = log(earnings_clean$earn)
earnings_clean$height_norm = earnings_clean$height_ctr/sd(earnings_clean$height)
earnings_clean$yearbn_ctr_decade = earnings_clean$yearbn_ctr/10
earnings_clean$log_earn = log(earnings_clean$earn)

earn_model_1a = lm(earn ~ height_norm + yearbn_ctr_decade + sex, data = earnings_clean)
earn_model_1b = lm(log_earn ~ height_norm + yearbn_ctr_decade + sex, data = earnings_clean)
# earn_model_1c = lm(log_earn ~ height_norm_sex + yearbn_ctr_decade + sex, data = earnings_clean)

par(mfrow = c(2, 2))
plot(earn_model_1a)
plot(earn_model_1b)

```

Model 2: going with the log_earn model because it somewhat controls the outlying residuals a little better.

```{r}
earn_model_2 = lm(log_earn ~ height_norm + yearbn_ctr_decade * sex, data = earnings_clean)

summary(earn_model_2)

```

Model 3: leaving in sex/yearbn interaction

```{r}
earn_model_3 = lm(log_earn ~ yearbn_ctr_decade * sex + height_norm * yearbn_ctr_decade, data = earnings_clean)

summary(earn_model_3)

earn_model_3b = lm(log_earn ~ sex + height_norm * yearbn_ctr_decade, data = earnings_clean)

summary(earn_model_3b)

```

It seems that the height\*year interaction might explain some of the sex\*interaction. So I removed the latter. The R-sq remains the same (11%). The model doesn't explain a lot of the variance and the diagnostics aren't great.

```{r}
par(mfrow = c(2, 2))

plot(earn_model_3b)
```

### c) Interpret the model coefficients:

Note: x times difference can be rescaled relative to 1 and interpretted as percentage. If you deal with risk differences (arithmetic differences in risk), this is confusing.

Intercept: The average earnings are $21K for a male of average population height (67 inches) born in 1949.

sexFemale: Earnings for females are about 0.66 ($e^{-0.41}$) times lower than a male of same height and birth year. I.e., the earnings change by -44% between males and females of same age and height.

height_norm: Earnings increase an average 1.1 ($e^{0.099}$) times per std. deviation increase in height between two people of the same sex born in 1949. 

yearbn_ctr_decade: Earnings decrease an average 0.91 ($e^{-0.085}$) times per decade increase in birth year between two people of the same sex who are 67 inches in height. 

height_norm:yearbn_ctr_decade: Between two people of the same sex, for each std. deviation increae in height, earnings decrease 0.91 ($e^{--0.06}$) times further for each decade increase in birth year. Between two people of the same sex, earnings increased per std. deviation increse in height are (modulated) decreased by 0.91 ($e^{--0.06}$) times per each decade increase in birth year. 

## Question 3 


The following was generated from a regression of weight on age per 10 years. The output was provided and curated by IamGianluca on github.

```{r, eval = F, include = T}
lm(formula = weight ~ age10)
                  coef.est coef.se 
(Intercept)       161.0     7.3 
age10             2.6       1.6
---
  n = 2009, k = 2
  residual sd = 119.7, R-Squared = 0.00


lm(formula = weight ~ age10 + age10.sq) 
                  coef.est coef.se
(Intercept)       96.2      19.3 
age10             33.6      8.7 
age10.sq          -3.2      0.9
---
  n = 2009, k = 3, 
  residual sd = 119.3, R-Squared = 0.01


lm(formula = weight ~ age30.44 + age45.64 + age65up)
                  coef.est coef.se
(Intercept)       157.2    5.4
age30.44TRUE      19.1     7.0
age45.64TRUE      27.2     7.6
age65upTRUE       8.5      8.7
---
  n = 2009, k = 4
  residual sd = 119.4, R-Squared = 0.01
```

### a-c) plot the regression results for each model.

(I'm assuming age ranges between 18 and 80)

```{r}
par(mfrow = c(1,1))
age = seq(18,80, by = 1)
weight = seq(140, 190, length = length(age))

plot(age, weight, "n")
curve(cbind(1, x/10) %*% c(161, 2.6), add = T, col = "red")
curve(cbind(1, x/10, x^2/100) %*% c(96.2, 33.6, -3.2), add = T, col = "blue")
#model 3
curve(cbind(1,x) %*% c(157.2, 0), add = T, xlim = c(18, 30))
curve(cbind(1,x) %*% c(157.2 + 19.1, 0), add = T, xlim = c(30, 44))
curve(cbind(1,x) %*% c(157.2 + 27.2, 0), add = T, xlim = c(44, 64))
curve(cbind(1,x) %*% c(157.2 + 8.5, 0), add = T, xlim = c(64, 80))
legend(x = 50, 160, c("model1", "model2", "model3"), lty = 1, col = c("red", "blue", "black"))

```

## Question 4

Using `pollution` data.

```{r}
pollution = foreign::read.dta("http://www.stat.columbia.edu/~gelman/arm/examples/pollution/pollution.dta")
```

### a) Plot and fit an lm of mortality rate versus nitric oxide. Show diagnostics.

```{r}
with(pollution, plot(nox, mort))
model4a = lm(mort ~ nox, data = pollution)
par(mfrow = c(2,2))
plot(model4a)
par(mfrow = c(1,1))
```

### b) Transform the data to adjust for issues seen in a)


log looks good for the nox variable. Using a log(predictor), choose a base that makes the interpretation works. Otherwise, Log the outcome and use the elasticity interpretation. Choosing the former.

```{r}
with(pollution, plot(log(nox), mort))
model4b1 = lm(mort ~ log10(nox), data = pollution)
model4b2 = lm(log(mort) ~ log(nox), data = pollution)

par(mfrow = c(2,2))
plot(model4b1)
plot(model4b2)
par(mfrow = c(1,1))

summary(model4b1)
#summary(model4b2)

```

### c) For every 10 times increase in nitric oxide, mortality rate increases 35.1. (log-log: per every unit increase in nitric acid, mortality rate increases about 1.5%)

### d) Add in sulfur dioxide and hydrocarbons

```{r}
par(mfrow = c(2,2))
with(pollution, {plot(so2, mort); plot(hc, mort)})
with(pollution, {plot(log10(so2), mort); plot(log10(hc), mort)})

model_4d = lm(mort ~ log10(nox) + log10(so2) + log10(hc), data = pollution)
summary(model_4d)
plot(model_4d)


```

Model seems ok (24% R-sq), there is an outlier (low so2).

### e) cross-validation, use model in d) to fit half of data then predict second half

```{r}
set.seed(15)
row_sample = sample(dim(pollution)[1], round(dim(pollution)[1]/2))

model_4e = lm(mort ~ log10(nox) + log10(so2) + log10(hc), data = pollution[row_sample, ])

fitted_values = predict(model_4e, pollution[-row_sample, ])

plot(fitted_values, pollution[-row_sample, "mort"], ylab = "cross-v data", main = "dotted line = 1sd")
abline(0, 1)
abline(summary(model_4e)$sigma, 1, lty = 2)
abline(-summary(model_4e)$sigma, 1, lty = 2)

```

## Question 6. Interpret the coefficient in the following model:

Model: $log(Q) = \alpha + \beta log(P)$.     
Q = smokes bought.    
P = cost of cigarettes.    
unit = counties in US.
$\beta = 0.3$

Per every unit increase in average price in cigarettes, the purchased quantity inceases by `r round(exp(0.3), 2)` ($e^{0.3}$) times. 


