---
title: "Chapter 3 exercises"
output: html_document
---

```{r data and packages, echo = F, message = F}
#library(RCurl)

example_data = as.data.frame(data.table::fread("ARM_Data/exercise2.1.dat"))

```

## Question 1

Fitting example data

### a) fit linear regression, summarize inference
### b) plot the results

```{r}
model1 = lm(y ~ x1 + x2, data = example_data[1:40, ])
summary(model1)
sigma = summary(model1)$sigma


with(example_data, plot(x1, y))
curve(cbind(1, x,  mean(example_data$x2[1:40])) %*% coef(model1), add = T)

with(example_data, plot(x2, y))
curve(cbind(1, mean(example_data$x1[1:40]), x) %*% coef(model1), add = T)

```

### c) plot residuals

```{r}

residuals = example_data$y[1:40] - predict(model1, example_data[1:40,])

with(example_data[1:40, ], plot(x1, residuals))
abline(a = 0, b = 0)
abline(a = -sigma, b = 0, lty = 2)
abline(a = sigma, b = 0, lty = 2)

with(example_data[1:40, ], plot(x2, residuals))
abline(a = 0, b = 0)
abline(a = -sigma, b = 0, lty = 2)
abline(a = sigma, b = 0, lty = 2)

```

### d) predict using the remaining 20 data points

The diagnostics aren't great but the predictions seem ok.

```{r}
predict_y = predict(model1, example_data[41:60, ])

with(example_data, plot(x1, y))
with(example_data[41:60,], points(x1, predict_y, pch = 2))

with(example_data, plot(x2, y))
with(example_data[41:60,], points(x2, predict_y, pch = 2))
```

## Question 2

log earnings ~ log height

* 66 inch tall person makes $30K

* 1% height increase = 0.8% increase in earnings

* 95% of predicted earnings fall within a factor of 1.1

### a) give the line and model sd

*First the equation*

Model 2: $Log E2 = log(\$30K) + m * log(H2/66)$    
Model 1: $log E1 = log(\$30K) + m * log(H1/66)$

If there is an 0.8% increase in earnings that means:
$$E2 = 1.008 * E1 $$
$$log(E2) - log(E1) \approx 0.008 $$

This is per 1% height increase so similarly:
$$log(H2) - log(H1) \approx 0.01 $$

So if you do model 2 - model 1:
$$log(E2) - log(E1) = m * (log(H2) - log(H1)) $$
$$0.008 = m * 0.01$$
$$m =  0.8 $$

The final equation is:
$$ Log(E) = log($30K) + 0.998 *log(H/66) $$

*now the sd (which I'm not entirely sure about and this is surely wrong)*

$$  log(1) +- 1.96 * log(sd) = log(1.1) $$
$$ sd = exp(log(1.1)/1.96) $$

### b) suppose sd of log height is 5%, what is R^2

The equation for R-sq: $R^2 = 1 - \sigma/{\sigma_y}^2$

But we are given $\sigma_x$.If we assume that is a typo. 5% tells us that for earnings, the sd is about 1.05. so we end up with:
$$ 1 - 1.049^2/1.05^2 = 0$$


## Question 3

### a) Generate 1000 N(0,1) for two variables and run the regression

```{r}
set.seed(5)
var1 = rnorm(1000)
var2 = rnorm(1000)

model3 = lm(var2 ~ var1)
summary(model3)


```

### b) repeat this 100 times, save z, how many z > 2
```{r}
set.seed(150)
z_tests = rep(NA, 100)
for(i in 1:100){
  var1 = rnorm(1000)
  var2 = rnorm(1000)
  fit = lm(var2 ~ var1)
  z_tests[i] = coef(fit)[2]/coef(summary(fit))[2,2]
}
hist(z_tests, breaks = 20)
mean(abs(z_tests) > 2)

```

## Question 4

Load the child iq data. it has age 3 IQ scores with moms age at childs birth and moms education for 400 ids.

```{r}
library(foreign)
iq.data = read.dta("ARM_Data/child.iq/child.iq.dta")
```

### a) Fit IQ on age, plot it, check assumptions and interpret the slope.

```{r}
model_q4 = lm(ppvt ~ momage, data = iq.data)
summary(model_q4) 

plot(iq.data$momage, model_q4$residuals);abline(0,0)

with(iq.data, plot(momage, ppvt))
curve(coef(model_q4)[1] + coef(model_q4)[2] * x, add = T)

```

Child's IQ increases 0.84 on average per year increase in mom's age at child's birth. The model fit is poor but the effect is signficant.

### b) Add in education.
```{r}
model_q4b = lm(ppvt ~ momage + educ_cat, data = iq.data)
summary(model_q4b) 

with(subset(iq.data, educ_cat == 1), plot(momage, ppvt))
curve(cbind(1, x, 1) %*% coef(model_q4b), add = T)
for(i in 2:4){
  with(subset(iq.data, educ_cat == i), points(momage, ppvt, col = i))
  curve(cbind(1, x, i) %*% coef(model_q4b), add = T, col = i)
}

```

For a given mom's age, IQ increases by 4.7 points per increase in education category. For a given education category, IQ increases by 0.34 per year increase in mom's age (not significant).

### c) Make a new variable for whether mom completes high school. Add an interaction.

```{r}
iq.data$mom_posths = with(iq.data, ifelse(educ_cat > 1, 1, 0))

model_q4c = lm(ppvt ~ momage * mom_posths, data = iq.data)
summary(model_q4c) 

with(subset(iq.data, mom_posths == 0), plot(momage, ppvt))
curve(cbind(1, x, 0, 0) %*% coef(model_q4c), add = T)

with(subset(iq.data, mom_posths == 1), points(momage, ppvt, col = 2))
curve(cbind(1, x, 1, x) %*% coef(model_q4c), add = T, col = 2)

```

### d) Repeat model b for first 200, predict next 200, compare
```{r}
train_data = iq.data[1:200, ]
model_q4d = lm(ppvt ~ momage + educ_cat, data = train_data)

predict_data = iq.data[201:400,]
predict_data$predict.iq = predict(model_q4d, predict_data)

with(predict_data, plot(y = ppvt, x = predict.iq, "n"))
for(i in 1:4){
  with(predict_data, 
     points(y = ppvt[educ_cat == i], x = predict.iq[educ_cat == i], col = i))
}
abline(0, 1)

```


## Question 5

Using the beauty data, student evaluation of instructors beauty and teaching quality for courses at UTexas. Evaluations come at end of course, beauty scores come from non-students (at a later time).


```{r}
beauty = readr::read_csv("http://www.stat.columbia.edu/~gelman/arm/examples/beauty/ProfEvaltnsBeautyPublic.csv")

hist(beauty$courseevaluation)
```

### a) fit a model testng if beauty predicts outcome.

```{r}
beauty_model = lm(courseevaluation ~ female + btystdave + age + minority + nonenglish + students + tenured, data = beauty)

summary(beauty_model)

par(mfrow = c(2,2))
plot(beauty_model)
par(mfrow = c(1,1))

colors <- ifelse(beauty$female == 1, "red", "blue")
plot(beauty$btystdave, beauty$courseevaluation, col = colors, pch = 20)
for(i in 0:1)
  curve(cbind(1, i, x, 48, 0.5, 0.5, 30, 0.5) %*% coef(beauty_model), 
        add = T,
        col = data.frame(x = c("blue", "red"), stringsAsFactors = F)$x[i + 1])
legend(1, 2.85, legend = c("Lady", "Dude"), col = unique(colors), pch = 20)

```

### b) mess around with other models

Based on my first model, I dumped a few variable that didn't seem important. I dichomitimized age. I standardardized student total. Tenure and tenure track have a lot of overlap, basically because tenured == 1 corresponds to tenuretrack. I threw an 3-way interaction in there for fun.


```{r}
beauty$old = ifelse(beauty$age > 50, 1, 0)
beauty$studentz = with(beauty, (log(students) - mean(students))/sd(students))
beauty$large_class = with(beauty, students > 50)


beauty_model_full = lm(courseevaluation ~ female * btystdave * old + nonenglish + large_class + tenured, data = beauty)

summary(beauty_model_full)

```

The model isnt that great ($R^2 = 16%$) but the diagnostics seem ok.

```{r}
par(mfrow = c(2,2))
plot(beauty_model_full)
par(mfrow = c(1,1))

```

```{r}
library(ggplot2)

prediction_data = expand.grid(
  btystdave = seq(-1.54, 1.88, by = 0.1),
  female = c(0, 1),
  old = c(0, 1),
  nonenglish = 0,
  large_class  = F,
  tenured = 0.5
)
prediction_data$courseevaluation = predict(beauty_model_full, prediction_data)

ggplot(data = subset(beauty, nonenglish == 0),
       aes(x = btystdave, y = courseevaluation, 
                          colour = factor(old), linetype = factor(old),
           shape = large_class)) +
  geom_point() +
  geom_line(data = prediction_data) +
  #geom_smooth(method= "lm") +
  facet_wrap(~female, labeller = 
               labeller(female = c("0" = "Dude", "1" = "Lady"),
             nonenglish = c("0" = "English-speaker", "1" = "Non-english-speaker"))) +
  theme_bw() +
  scale_color_discrete("Age", labels = c("<= 50", "> 50")) +
#  scale_shape_discrete("Non-english", labels = c("No", "Yes")) +
  scale_linetype_discrete("Age", labels = c("<= 50", "> 50")) +
  theme(legend.position = "top", legend.box = "horizontal")

```

There are repeat teachers, so lets average their scores.

```{r}
library(dplyr)
beauty_agg = beauty %>% 
  group_by(profnumber, btystdave, female, old, nonenglish, tenured) %>%
  summarize(
    avg_course_rating = mean(courseevaluation),
    pct_large_class = mean(large_class),
    large_class_majority = 1 * (pct_large_class > 0.5)
  )


beauty_model2 = lm(avg_course_rating ~ female * btystdave * old + nonenglish + large_class_majority + tenured, data = beauty_agg)

summary(beauty_model2)

par(mfrow = c(2,2))
plot(beauty_model2)
par(mfrow = c(1,1))
```

```{r}
prediction_data2 = expand.grid(
  btystdave = seq(-1.54, 1.88, by = 0.1),
  female = c(0, 1),
  old = c(0, 1),
  nonenglish = 0,
  large_class_majority  = c(0,1),
  tenured = 0.5
)
prediction_data2$avg_course_rating = predict(beauty_model2, prediction_data2)

ggplot(data = beauty_agg,
       aes(x = btystdave, y = avg_course_rating, 
                          colour = factor(old), linetype = factor(old),
           shape = factor(nonenglish))) +
  geom_point() +
  geom_line(data = prediction_data2) +
  facet_grid(large_class_majority~female, labeller = 
               labeller(
                 female = c("0" = "Dude", "1" = "Lady"),
                 large_class_majority = 
                    c("0" = "Most classes < 50 students", 
                     "1" = "Most classes > 50 students"),       
                 nonenglish = c("0" = "English-speaker", 
                                "1" = "Non-english-speaker")),
             ) +
  theme_bw() +
  scale_color_discrete("Age", labels = c("<= 50", "> 50")) +
  scale_shape_discrete("Non-english", labels = c("No", "Yes")) +
  scale_linetype_discrete("Age", labels = c("<= 50", "> 50")) +
  theme(legend.position = "top", legend.box = "horizontal")

```