---
title: "Solutions to some computer exercises in the BDA book by Gelman et al. (2nd ed.)"
author: "Bryan Mayer"
date: "January 8, 2017"
output: pdf_document
---

```{r setup, include=FALSE}

library(ggplot2)
library(dplyr)
library(knitr)
library(xtable)

load('BDA-data.RData')

```

# Chapter 2

## Exercise 2.13

### Data:
```{r, echo = F, results = "asis"}
airline_data %>% xtable(caption = "Airline data from Table 2.2 (BDA 2nd ed.). Death rate is passenger deaths per 100 million passenger miles \\label{tab:airline}") %>% 
  print(comment = F, include.rownames = F)

```

### (a) Assume accidents are independent with Poisson($\theta$). Set a prior for $\theta$ and find posterior distribution using the data. Give a 95% predictive interval for fatal accidents in 1986.

Solution: Use gamma(0,0) as a non-informative prior. Conjugate posterior is gamma(0 + 10 * mean(y), 0 + 10)

```{r}
sim_theta = rgamma(5000, shape = 10 * mean(airline_data$Fatal_accidents), rate = 10)
hist(sim_theta, xlab = "theta|y")
summary(sim_theta)
quantile(sim_theta, c(0.025, 0.975))

predictive_post = rpois(5000 , sim_theta)
hist(predictive_post, xlab = "y~|theta")
quantile(predictive_post, c(0.025, 0.975))

```

### (b) Assume accidents each year follow independent Poisson with constant rate times numbers of passenger miles flown. Set a prior distribution for theta and repeat analysis of (a). For the predictive distribution, assume 8e11 passenger miles were flown in 1986.

Solution: Exposure variable is 1/death rate (100 million passenger miles/death) * passenger deaths * 1e8

likelihood is conditioned on exposure variable too, so 
$$
y|\theta, E \sim Poisson(\theta * E)
$$

The posterior distribution is now gamma(0 + 10$\bar{y}$, 0 + 10$\bar{E}$).


```{r}
exposure_var = airline_data$Passenger_deaths/airline_data$Death_rate * 1e8

sim_theta = rgamma(5000, shape = 10 *mean(airline_data$Fatal_accidents), rate = 10 * mean(exposure_var))
hist(sim_theta, xlab = "theta|y")
summary(sim_theta)
quantile(sim_theta, c(0.025, 0.975))

predictive_post = rpois(5000 , sim_theta * 8e11)
hist(predictive_post, xlab = "y~|theta")
quantile(predictive_post, c(0.025, 0.975))

```